---
title: "Final Report: [Obesity]"
date: 2026-06-04
author: "Anna Usoltseva"

format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 4
---

# Final Report: \[Obesity\]

## Introduction

Obesity has emerged as a major public health problem. Centers for Disease Control and Prevention (CDC) reported the prevalence rate of obesity in the United States as 40.3% during 2021 - 2023 period \[1\]. Unlike acute illnesses, the health consequences of obesity typically develop gradually. Obesity accelerates biological aging by promoting chronic low-grade inflammation, oxidative stress, and mitochondrial dysfunction, all of which contribute to the earlier onset of age-related diseases \[2\]. It increases the risk of numerous chronic conditions, including type 2 diabetes mellitus, cardiovascular disease, metabolic syndrome, hypertension, and certain types of cancers \[3, 4\]. In this project I want to test the relationship between lifestyle factors and obesity. I expect for example that regular physical activity is associated with lower obesity levels. Hypothesis (H1): Healthier lifestyle habits are associated with lower obesity rates. Null hypothesis (H0): There is no correlation between lifestyle and obesity. The findings may contribute to a better understanding of the problem and alter public health strategies.

## Methods

Obesity Prediction dataset was used in this project. It is available on Kaggle:
https://www.kaggle.com/datasets/ruchikakumbhar/obesity-prediction 

The dataset included 2,111 participants and 17 variables related to demographics, lifestyle, and obesity level.

## Results

### Library

```{r}
#| warning: false

# Library
library(ggplot2)
library(tidyverse)
library(corrplot)
library(GGally)
library(car)
library(rstatix)
```

### Structure of the dataset, summary statistics, missing values, and the distribution of obesity levels. 

```{r}
getwd()
setwd("/Users/kulkins/Desktop/R_s/AnnaUsoltseva.github.io/projects")
list.files()
Obesity <- read.csv("Obesity prediction.csv")
glimpse(Obesity)
names(Obesity)
summary(Obesity)
colSums(is.na(Obesity))
table(Obesity$Obesity)

# The obesity levels were originally ordered alphabetically, so they were reordered into the ordinal sequence

Obesity$Obesity <- factor(
  Obesity$Obesity,
  levels = c(
    "Insufficient_Weight",
    "Normal_Weight",
    "Overweight_Level_I",
    "Overweight_Level_II",
    "Obesity_Type_I",
    "Obesity_Type_II",
    "Obesity_Type_III"))


levels(Obesity$Obesity)
table(Obesity$Obesity)

# Visualization. How many human subjects was in each  group?

ggplot(Obesity, aes(x = Obesity)) +
  geom_bar(fill = "skyblue") +
  geom_text(stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.3,
    size = 5) +
  labs(title = "Distribution of Obesity Levels",
    x = "Obesity Level",
    y = "Number of Participants") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1))
```

### FAMILY HISTORY.

```{r}
table(Obesity$family_history)
summary(Obesity$family_history)

table_history <- table(Obesity$family_history, Obesity$Obesity)
table_history
```


#### How did family history of obesity vary across the different obesity levels?

```{r}
chisq.test(table_history)

ggplot(Obesity, aes(x = Obesity, fill = family_history)) +
  geom_bar(position = "dodge") +
  geom_text(stat = "count",
    aes(label = after_stat(count)),
    position = position_dodge(width = 0.9),
    vjust = -0.3,
    size = 3.5) +
  scale_fill_manual(values = c(
      "no" = "skyblue",
      "yes" = "indianred")) +
  labs(title = "Obesity Level and Family History",
    x = "Obesity Level",
    y = "Number of Participants",
    fill = "Family History") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1))

# The Chi-square test showed a significant association between family history and obesity level (χ² = 621.98, p < 0.001). Participants with a family history of obesity were much more common in the higher obesity categories.
```

### PHYSICAL ACTIVITY. Was the physical activity data normally distributed?

```{r}
summary(Obesity$FAF)

hist(Obesity$FAF,
     breaks = 10,
  col = "skyblue",
  border = "black",
  main = "Histogram of Physical Activity (FAF)",
  xlab = "Physical Activity (FAF)")


ggplot(Obesity, aes(sample = FAF)) +
  stat_qq() +
  stat_qq_line(color = "red", linewidth = 1) +
  labs(
    title = "Q-Q Plot of Physical Activity (FAF)",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles") +
  theme_classic()

shapiro.test(Obesity$FAF)

# Was the assumption of homogeneity of variance met?

leveneTest(FAF ~ Obesity, data = Obesity)

#Physical activity was not normally distributed, and the assumption of equal variances was also violated. Therefore, non-parametric methods were used for the analysis. 
```

#### How do the physical activity levels compare across obesity levels?

```{r}
# Was there a significant difference in physical activity across obesity levels?

kruskal.test(FAF ~ Obesity,
             data = Obesity)

# Were there significant pairwise differences in physical activity among obesity levels?

games_howell_test(
  data = Obesity,
  FAF ~ Obesity) %>%
print(n = Inf, width = Inf)

gh_results <- games_howell_test(
  data = Obesity,
  FAF ~ Obesity)

view(gh_results)

# Summary Table of Physical Activity by Obesity Level
summary_df <- Obesity %>%
  group_by(Obesity) %>%
  summarise(Mean = mean(FAF),
    SD = sd(FAF),
    n = n(),
    SE = SD / sqrt(n),
    CI = qt(0.975, n - 1) * SE,
    .groups = "drop")
summary_df

ggplot(summary_df, aes(x = Obesity, y = Mean)) +
  geom_col(fill = "lightgreen", width = 0.7) +
  geom_errorbar(
    aes(
      ymin = Mean - CI,
      ymax = Mean + CI),
    width = 0.2,
    linewidth = 0.8) +
  labs(
    title = "Physical Activity Across Obesity Levels",
    x = "Obesity Level",
    y = "Mean Physical Activity (FAF)"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1))

# People with Insufficient Weight and Normal Weight had the highest average physical activity, while participants with Obesity Type III had the lowest average physical activity. Several pairwise comparisons were statistically significant according to the Games-Howell test. 
```


#### How do physical activity levels compare across obesity levels and between genders?

```{r}
#| warning: false
# Gender differences
Obesity %>%
  count(Obesity, Gender)

# Summary Table for Gender Differences
summary_gender_plot <- Obesity %>%
  group_by(Obesity, Gender) %>%
  summarise(Mean = mean(FAF, na.rm = TRUE),
    SD = sd(FAF, na.rm = TRUE),
    n = sum(!is.na(FAF)),
    SE = SD / sqrt(n),
    CI = qt(0.975, df = n - 1) * SE,
    .groups = "drop") %>%
  filter(n >= 10) %>%
  droplevels()

print(summary_gender_plot)

gender_plot <- ggplot(summary_gender_plot,
  aes(x = Obesity,
    y = Mean,
    fill = Gender)) +
  geom_col(width = 0.8) + geom_errorbar(
    aes(ymin = pmax(0, Mean - CI),
      ymax = Mean + CI),
    width = 0.2,
    linewidth = 0.8) +

  facet_grid(cols = vars(Gender),
    scales = "free_x",
    space = "free_x") +

  scale_fill_manual(
    values = c(
      "Female" = "hotpink",
      "Male" = "steelblue")) +

  scale_x_discrete(drop = TRUE,
    labels = c(
      "Insufficient_Weight" = "Insufficient",
      "Normal_Weight" = "Normal",
      "Overweight_Level_I" = "Overweight I",
      "Overweight_Level_II" = "Overweight II",
      "Obesity_Type_I" = "Obesity I",
      "Obesity_Type_II" = "Obesity II",
      "Obesity_Type_III" = "Obesity III")) +
scale_y_continuous(breaks = seq(0, 2, by = 0.5),
    expand = expansion(mult = c(0, 0.03))) +
coord_cartesian(ylim = c(0, 2)) + labs(
    title = "Mean Physical Activity Across Obesity Levels by Gender",
    x = "Obesity Level",
    y = "Mean Physical Activity (FAF)") +

  theme_classic() +

  theme(legend.position = "none",
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"),
    strip.text = element_text(
      face = "bold",
      size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1))

print(gender_plot)

# Males generally had higher average physical activity than females within the same obesity category. Some obesity categories contained very few participants of one gender, and was excluded from the plot.
```


#### How do physical activity levels vary across obesity levels and family history groups?

```{r}
# 

heat_data <- Obesity %>%
  group_by(Obesity, family_history) %>%
  summarise(mean_FAF = mean(FAF),
    n = n(),
    .groups = "drop") %>%
  filter(n >= 10)


ggplot(heat_data,
       aes(x = family_history,
           y = Obesity,
           fill = mean_FAF)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(
      round(mean_FAF, 2),
      "\n(n = ", n, ")")),
    size = 4) +
  scale_fill_gradient(
    low = "lightcyan",
    high = "steelblue") +
  labs(title = "Average Physical Activity by Obesity Level and Family History",
    x = "Family History",
    y = "Obesity Level",
    fill = "Mean FAF") +
  theme_classic()

# The heat map showed that mean physical activity generally decreased as obesity level increased, especially among participants with a family history of obesity.
```


#### How does physical activity vary with age across obesity levels?

```{r}
summary(Obesity$Age)

ggplot(Obesity, aes(Age, FAF)) +
  geom_jitter(width = 0.25, height = 0.05, alpha = 0.35) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  facet_wrap(~Obesity, ncol = 2) +
  theme_classic()

# The scatter plots showed a weak relationship between age and physical activity across most obesity groups. However, the majority of participants were between 20 and 40 years of age, resulting in a limited sample size for older age groups. Therefore, conclusions regarding older participants should be interpreted with caution. The strongest age-related decline in physical activity was observed among participants with Obesity Type III.
```


### HIGH-CALORIE FOODS INTAKE (FAVC). Is the consumption of high-calorie foods (FAVC) associated with obesity level?

```{r}
summary(Obesity$FAVC)

table_FAVC <- table(
  Obesity$FAVC,
  Obesity$Obesity)

table_FAVC

chisq.test(table_FAVC)

count_data <- Obesity %>%
  count(Obesity, FAVC) %>%
  mutate(x_num = as.numeric(Obesity))

ggplot(count_data,
  aes(x = x_num, y = n, fill = FAVC)) +
  geom_col(position = position_dodge(width = 0.9),
    width = 0.8) +
  geom_smooth(aes(color = FAVC,
      group = FAVC),
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    linewidth = 1.2,
    show.legend = FALSE) +
  geom_text(aes(label = n),
    position = position_dodge(width = 0.9),
    vjust = -0.3,
    size = 3.5) +
  scale_x_continuous(breaks = 1:7,
    labels = levels(Obesity$Obesity)) +
  scale_fill_manual(values = c(
      "no" = "skyblue",
      "yes" = "indianred")) +
  scale_color_manual(values = c(
      "no" = "steelblue",
      "yes" = "red3")) +
  labs(title = "Obesity Level and High-Calorie Food Intake",
    x = "Obesity Level",
    y = "Number of Participants",
    fill = "High-Calorie Intake") +
  theme_classic() +
  theme(plot.title = element_text(
      hjust = 0.5,
      face = "bold"),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1))

# The Chi-square test also found a significant association between obesity and consumption of high-calorie foods (χ² = 233.34, p < 0.001). High-calorie food intake was more common among participants in the obesity categories.
```



## Conclusions

The results suggest that obesity is influenced by several factors. Participants with higher obesity levels generally had lower physical activity, were more likely to report a family history of obesity, and more frequently consumed high-calorie foods.

Limitation of this study is that the dataset is observational, so it cannot prove cause and effect. In addition, some obesity groups contained very few participants of one gender, making those comparisons less reliable.

## References:

1.  National Center for Health Statistics (2024) Obesity and severe obesity prevalence in adults: United States, August 2021–August 2023 (Data Brief No. 508). Hyattsville, MD: Centers for Disease Control and Prevention.
2.  Salvestrini V, Sell C, Lorenzini A (2019) Obesity may accelerate the aging process. Front Endocrinol (Lausanne) 10, 266.
3.  Jin X, Qiu T, Li L, Yu R, Chen X, Li C, et al. (2023) Pathophysiology of obesity and its associated diseases. Acta Pharm Sin B 13, 2403-2424.
4.  Scully T, Ettela A, LeRoith D, Gallagher EJ (2021) Obesity, type 2 diabetes, and cancer risk. Front Oncol 10, 615375. Dataset: Obesity Prediction https://www.kaggle.com/datasets/ruchikakumbhar/obesity-prediction


------------------------------------------------------------------------

**Last updated:** June 2026
